---
title: "Cascade Gateway - explore"
output: html_notebook
---

The [Cascade Gateway Border Data Warehouse](http://www.cascadegatewaydata.com/Crossing) provides detailed, real-time data for the major Canada/U.S. border crossings between British Columbia and Washington in the Lower Mainland/Pacific Northwest.

* Lynden/Aldergrove

* Pacific Highway

* Peace Arch

* Sumas/Huntingdon

This is an initial exploration of the data, to understand the variables offered and how we might tabulate and visualize them.

Exploratory data analysis!

```{r setup}
# tidyverse packages
library(tidyverse)

# utilities
library(here)

```

## Data import, #1

A csv file was downloaded via the "Custom Query" menu, with the following specifications:

* Date: 2019-01-01 to 2020-03-31 (note the website uses mdy format)

* Detector: none selected

* Crossing: Peach Arch / Northbound / Peace Arch North Cars & Peace Arch North NEXUS

* Weigh-in-Motion (WIM): none selected

The file was renamed "peacearch_N_cars_20190101_2020-03-31.csv"

```{r}

df_peacearch <- read.csv(here("data", "peacearch_N_cars_20190101_2020-03-31.csv"))

df_peacearch 

```



