---
title: "Exploring Border Crossing API"
output: html_notebook
---

# Getting data from the API

Libraries/setup
```{r}
#install.packages("httr")

library(httr)
library(lubridate)
library(tidyverse)

options(stringsAsFactors = FALSE)
```


Setup the URL and path for the API call. Anything after the ? in path will be the query from the API. Here format = CSV
```{r}
url <- "http://www.cascadegatewaydata.com"
path <- "Crossing/?format=CSV"
raw.result <- GET(url = url, path= path)
```


For easier to read code, you can use the query argument:
```{r}
url <- "http://www.cascadegatewaydata.com"
path <- "Crossing"
query = list(format = "CSV")
raw.result <- GET(url = url, path= path, query = query)
```


Exploring result:
```{r}
names(raw.result)
```


Look at status_code. Result of 200 tells us the server received the request
```{r}
raw.result$status_code
```


For more detailed messaging use http_status
```{r}
http_status(raw.result)
```


Look at the content. It is unicode and not useful
```{r}
raw.result$content
```


Instead use content function
```{r}
content(raw.result)
```


# Exploring queries

Recreating custom query data "peacearch_N_cars_20190101_2020-03-31.csv"

* Date: 2019-01-01 to 2020-03-31 (note the website uses mdy format)

* Detector: none selected

* Crossing: Peach Arch / Northbound / Peace Arch North Cars & Peace Arch North NEXUS

* Weigh-in-Motion (WIM): none selected

Note that the query only lets you select one lane type at a time. Here only car is selected.

```{r}

url <- "http://www.cascadegatewaydata.com"
path <- "Crossing"
query <- list(id=134, start="01/01/2019", end="03/31/2020", data="avg-delay,sum-vol",
              dir="Northbound",lane="car",dg = "Day",tg="Day", format="CSV" )

raw <- GET(url = url, path = path, query = query) 
http_status(raw)
```

```{r}
peacearch_N_cars <- content(raw)
peacearch_N_cars %>% head(5)
```

## Options for selecting aggregated data

Note that some measurement of time is needed, without it the results return only the valid crossing IDs and their attributes.
```{r}
query <- list(id=134, data="avg-delay,sum-vol",
              dir="Northbound",lane="car", format="CSV" )

peacearch_N_cars_test <- GET(url = url, path = path, query = query) %>%
  content()

peacearch_N_cars_test
```


Year - Returns results aggregated by month. Ignores any group by values. Can only select one year at a time
```{r}
query <- list(id=134, year="2019", data="avg-delay,sum-vol",
              dir="Northbound",lane="car", format="CSV" )

peacearch_N_cars_test <- GET(url = url, path = path, query = query) %>%
  content()

peacearch_N_cars_test
```


Month - When selecting data by Year (above), can filter to receive results for only one month. This returns results aggregated by day. Can select only one month.
```{r}
query <- list(id=134, year = "2019", month = "3", data="avg-delay,sum-vol",
              dir="Northbound", lane="car", format="CSV" )

peacearch_N_cars_test <- GET(url = url, path = path, query = query) %>%
  content()

peacearch_N_cars_test
```

Day - When selecting data by Year and filtering by month (above), can filter to receive results for only one day. Returns results aggregated by 5 minutes. Can only select one day.
```{r}
query <- list(id=134, year = "2019", month = "3", day = "5", data="avg-delay,sum-vol",
              dir="Northbound", lane="car", format="CSV" )

peacearch_N_cars_test <- GET(url = url, path = path, query = query) %>%
  content()

peacearch_N_cars_test
```

Start/end date - When providing a start date, an end date also needs to be provided. To get up-to-date results, could link end date to sys.date but this does not work in an app to update the data. Here, using yesterday as the end date to get a full days worth of data. Aside: does lubridate function today() work on servers? Can pass it a timezone.
```{r}

end_date <- Sys.Date()-1
end_date_char <- paste0(month(end_date), "/", day(end_date), "/", year(end_date))

query <- list(id=134, start="01/01/2019", end=end_date_char, data="avg-delay,sum-vol",
              dir="Northbound",lane="car",dg = "Day",tg="Day", format="CSV" )
# 
peacearch_N_cars <- GET(url = url, path = path, query = query) %>% content()
peacearch_N_cars
```

## Data manipulation

Add in some date indicators (Year, Week of the Year, Day of the Week)
```{r}
peacearch_N_cars_dates <- peacearch_N_cars %>%
mutate(Week = week(`Group Starts`),
       Year = year(`Group Starts`),
       DayOfWeek = wday(`Group Starts`))
```


Summarize by week of the year (exclude partial weeks). Note that start of week is Jan 1 and not necessarily Sunday/Monday, etc.
```{r}
peacearch_N_cars_byweek <- peacearch_N_cars_dates %>%
  group_by(Year, Week) %>%
  mutate(First = min(`Group Starts`),
         Last = max(`Group Starts`)) %>%
  group_by(Year, Week, First, Last) %>%
  summarize(NumDays = n(),
            Volume = sum(`Sum - Volume (Peace Arch)`, na.rm = TRUE)) %>%
  filter(NumDays == 7) %>% ungroup()

peacearch_N_cars_byweek
```

Chart data by week of the year
```{r}
colours <- c(`2019` = "black", `2020` = "blue")

peacearch_N_cars_byweek %>%
  ggplot(aes(x=Week, y = Volume, fill=as.character(Year))) +
  geom_bar(stat= "identity", position = "identity", aes(alpha = 0.3)) +
  scale_fill_manual(values = colours) +
  theme_classic() +
  scale_x_continuous(expand= c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  guides(alpha = "none",
         fill = guide_legend(title = "Year")) +
  labs(title ="Northbound Car Border Crossings through the Peace Arch by Week")
  
```



Write a function to generate charts by lane type, summarizing all northbound border crossings

```{r}
query_by_crossing_id_and_lane <- function(id, lane) {
  
  url <- "http://www.cascadegatewaydata.com"
  path <- "Crossing/"
  query <- list(id=id, start="01/01/2019", end=end_date_char, data="sum-vol",
              dir="Northbound",lane=lane, dg = "Day",tg="Day", format="CSV" )
  
  data <- GET(url = url, path = path, query = query) %>% content()
  
  data
  
}

chart_volume_by_week <- function(lane){

  ids <- c(134,135,136,137)

  data <- purrr::map_dfc(ids, query_by_crossing_id_and_lane, lane = lane) %>%
    group_by(`Group Starts`) %>%
    summarize(sum = sum(`Sum - Volume (Peace Arch)`, 
                      `Sum - Volume (Pacific Highway)`, 
                      `Sum - Volume (Lynden/Aldergrove)`, 
                      `Sum - Volume (Sumas/Huntingdon)`, na.rm = TRUE)) %>%
    mutate(Week = week(`Group Starts`),
           Year = year(`Group Starts`),   
           DayOfWeek = wday(`Group Starts`)) %>%
    group_by(Year, Week) %>%
    mutate(First = paste(month(min(`Group Starts`), label = TRUE, abbr = TRUE), day(min(`Group Starts`)))) %>%
    group_by(Year, Week, First) %>%
    summarize(NumDays = n(),
              Volume = sum(sum, na.rm = TRUE)) %>%
    filter(NumDays == 7) %>% ungroup()

  
  ## Chart attributes
  colours <- c(`2019` = "black", `2020` = "blue")
  
  ## Date labels to better reference volumes
  ## Dates based on 2019 and are 1 day off for 2020 after Feb 29...
  x_labs <- data %>%
    filter(Year == 2019, Week %in% seq(1, 52, 4)) %>%
    pull(First) 
  
  chart <- data %>%
    ggplot(aes(x=Week, y = Volume, fill=as.character(Year))) +
    geom_bar(stat= "identity", position = "identity", aes(alpha = 0.3)) +
    scale_fill_manual(values = colours) +
    theme_classic() +
    scale_x_continuous(expand= c(0,0), breaks = seq(1, 52, 4), labels = x_labs) +
    scale_y_continuous(expand = c(0,0)) +
    guides(alpha = "none",
           fill = guide_legend(title = "Year")) +
    labs(title =paste("Northbound", lane ,"Border Crossings by Week"))

  chart
}
```


```{r message = FALSE}
car_chart <- chart_volume_by_week("Car")
truck_chart <- chart_volume_by_week("Truck")
bus_chart <- chart_volume_by_week("Bus")
fast_chart <- chart_volume_by_week("Fast")
nexus_chart <- chart_volume_by_week("Nexus")
```

```{r}
car_chart
truck_chart
bus_chart
fast_chart
nexus_chart
```
















