---
title: "Exploring Border Crossing API"
output: html_notebook
---

# Getting data from the API

Libraries/setup
```{r}
#install.packages("httr")

library(httr)
library(lubridate)
library(tidyverse)

options(stringsAsFactors = FALSE)
```


Setup the URL and path for the API call. Anything after the ? in path will be the query from the API. Here format = CSV
```{r}
url <- "http://www.cascadegatewaydata.com"
path <- "Crossing/?format=CSV"
raw.result <- GET(url = url, path= path)
```


For easier to read code, you can use the query argument:
```{r}
url <- "http://www.cascadegatewaydata.com"
path <- "Crossing"
query = list(format = "CSV")
raw.result <- GET(url = url, path= path, query = query)
```


Exploring result:
```{r}
names(raw.result)
```


Look at status_code. Result of 200 tells us the server received the request
```{r}
raw.result$status_code
```


For more detailed messaging use http_status
```{r}
http_status(raw.result)
```


Look at the content. It is unicode and not useful
```{r}
raw.result$content
```


Instead use content function
```{r}
content(raw.result)
```


***

# Exploring queries

Recreating custom query data "peacearch_N_cars_20190101_2020-03-31.csv"

* Date: 2019-01-01 to 2020-03-31 (note the website uses mdy format)

* Detector: none selected

* Crossing: Peach Arch / Northbound / Peace Arch North Cars & Peace Arch North NEXUS

* Weigh-in-Motion (WIM): none selected

Note that the query only lets you select one lane type at a time. Here only car is selected.

```{r}

url <- "http://www.cascadegatewaydata.com"
path <- "Crossing"
query <- list(id=134, start="01/01/2019", end="03/31/2020", data="avg-delay,sum-vol",
              dir="Northbound",lane="car",dg = "Day",tg="Day", format="CSV" )

raw <- GET(url = url, path = path, query = query) 
http_status(raw)
```

```{r}
peacearch_N_cars <- content(raw)
peacearch_N_cars %>% head(5)
```


